---
output: github_document
---

```{r}
#| include = FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clover

<!-- badges: start -->
[![R-CMD-check](https://github.com/rnabioco/clover/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/rnabioco/clover/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

clover facilitates analysis and visualization of nanopore tRNA sequencing
data, including differential expression, base-calling error analysis, and
modification co-occurrence networks.

**clover is under active development.** *Caveat emptor*.

## Installation

You can install the development version of clover from
[GitHub](https://github.com/rnabioco/clover) with:

``` r
# install.packages("pak")
pak::pak("rnabioco/clover")
```

## Example

clover reads output from the
[aa-tRNA-seq-pipeline](https://github.com/rnabioco/aa-tRNA-seq-pipeline)
and stores the results in a `SummarizedExperiment`.

```{r}
#| label: example
#| warning: false
#| message: false
library(clover)
library(SummarizedExperiment)

# Load pipeline results into a SummarizedExperiment
se <- create_clover(
  config_path = clover_example("ecoli/config.yaml"),
  sample_info = data.frame(
    sample_id = c(
      "wt-15-ctl-01",
      "wt-15-ctl-02",
      "wt-15-ctl-03",
      "wt-15-inf-01",
      "wt-15-inf-02",
      "wt-15-inf-03"
    ),
    condition = rep(c("control", "infected"), each = 3)
  )
)

# Differential tRNA abundance with DESeq2
dds <- run_deseq(
  assay(se, "counts"),
  colData(se),
  design = ~condition
)

res <- tidy_deseq_results(
  dds,
  contrast = c("condition", "infected", "control")
)

plot_volcano(res)
```

See `vignette("clover")` for a complete walkthrough.

## Related work

- [aa-tRNA-seq-pipeline](https://github.com/rnabioco/aa-tRNA-seq-pipeline) is the Snakemake pipeline that generates the data clover analyzes.
- [R2easyR](https://github.com/JPSieg/R2easyR) visualizes structure probing signals on RNA secondary structure diagrams.
- [nanoblot](https://github.com/SamDeMario-lab/NanoBlot) facilitates visualization of nanopore sequencing data, including a "virtual gel" plot.
