---
format: gfm
knitr:
  opts_chunk:
    collapse: TRUE
    comment: "#>"
    fig.path: "man/figures/README-"
    out.width: "100%"
---

<!-- README.md is generated from README.qmd. Please edit that file -->

# clover

<!-- badges: start -->
[![R-CMD-check](https://github.com/rnabioco/clover/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/rnabioco/clover/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

clover facilitates analysis and plotting of nanopore tRNA sequencing data.

🚧 **clover is under active development.** Caveat emptor. 🚧 

## Installation

You can install the development version of clover from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("rnabioco/clover")
```

## Example

clover uses a `SummarizedExperiment::RangedSummarizedExperiment()` to store and compute on data associated with nanopore tRNA sequencing experiments.

Here's an example of how to create the object, which includes:

- `fa`: a FASTA file containing the tRNA reference sequences.
- `pod5`: a list of pod5 files, one per experiment. Merge pod5 files with the [pod5 tool](https://github.com/nanoporetech/pod5-file-format) and `pod5 merge *.pod5` 
- `bam`: a list of BAM files, one per experimetnt, each containing **reads aligned** to the tRNA FASTA reference.
- `mod_bed` [*optional*]: a [BED](https://genome.ucsc.edu/FAQ/FAQformat.html#format1) file containing sites of tRNA modification.

```{r}
#| label: ex-clover-rse
#| eval: false
library(clover)

rse_clover <- create_clover(
  pod5    = list(
    "grande" = clover_data("grande.pod5"),
    "petite" = clover_data("petite.pod5")
  ),
  bam     = list(
    "grande" = clover_data("grande.bam"),
    "petite" = clover_data("petite.bam")
  ),
  fa      = clover_data("yeast.trna.fa.gz"),
  mod_bed = clover_data("yeast.trna.mods.bed.gz")
)

rse_clover
```

## Summary statistics

The object contains data in the following slots:

- `exp`: expression levels, per-tRNA.
- `berror`: base-calling error rates, per-tRNA and per-position.

## Plots

clover provides plotting functions to visualize tRNA-related data.

The plots below illustrate the base-calling error rates and secondary structure of a tRNA molecule in the **grande** *S. cerevisiae* strain. The error rates are shown as a heatmap, with the x-axis representing the position in the tRNA molecule and the y-axis representing the tRNA molecule itself.

```{r}
#| label: fig-clover-heatmap
#| eval: false
plot_bcerror_heatmap(rse_clover)
```

The secondary structure is shown as a dot-bracket diagram, with the nucleotides colored by their base-calling error rates

```{r}
#| label: fig-clover-roadkill
#| eval: false
plot_trna_structure(
  rse_clover,
  exp = "grande",
  ref = "tRNA-Gly-GCC-1-1"
)
```

## Comparison

clover also provides functions to compare two `RangedSummarizedExperiment` objects corresponding to two experiments.

Here we calculate differential expression and modification of tRNAs prepared from **grande** and **petite** *S. cerevisiae* strains. The **petite** strain lacks mitochondrial DNA and its encoded tRNAs.

```{r}
#| label: ex-clover-comparison
#| eval: false
calc_diff_exp(rse_clover, ref = "grande", exp = "petite")

calc_diff_mod(rse_clover, ref = "grande", exp = "petite")
```

## Related work

- [R2easyR](https://github.com/JPSieg/R2easyR) vizualizes structure probing signals on RNA secondary structure diagrams.
- [nanoblot](https://github.com/SamDeMario-lab/NanoBlot) faciliates visualization nanopore sequencing data, including a "virtual gel" plot.
