---
title: "Introduction to clover"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to clover}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  warning = FALSE
)
```

```{r}
#| label: setup
#| message: false
library(clover)
library(dplyr)
library(tidyr)
```

## Overview

This vignette demonstrates clover's analysis workflow using *E. coli* tRNA
sequencing data from a T4 phage infection time-course experiment. The test
dataset includes 6 samples: 3 uninfected controls (`ctl`) and 3 T4-infected
(`inf`) wild-type *E. coli* at 15 minutes post-infection, with 3 biological
replicates per condition.

## Loading data with `create_clover()`

The simplest way to load results from the
[aa-tRNA-seq-pipeline](https://github.com/rnabioco/aa-tRNA-seq-pipeline)
is with `create_clover()`, which reads a pipeline configuration file and
assembles a `SummarizedExperiment`.

```{r}
#| label: create-se
config_path <- clover_example("ecoli/config.yaml")

sample_info <- data.frame(
  sample_id = c(
    "wt-15-ctl-01",
    "wt-15-ctl-02",
    "wt-15-ctl-03",
    "wt-15-inf-01",
    "wt-15-inf-02",
    "wt-15-inf-03"
  ),
  condition = rep(c("ctl", "inf"), each = 3),
  replicate = rep(1:3, 2)
)

se <- create_clover(
  config_path,
  types = c("charging", "bcerror", "odds_ratios"),
  sample_info = sample_info
)

se
```

The SE object contains:

- **assay "counts"**: abundance count matrix (charged + uncharged reads)
- **colData**: sample metadata with condition labels
- **metadata**: list with raw charging data, bcerror, odds ratios, FASTA reference, and pipeline config

```{r}
#| label: inspect-se
SummarizedExperiment::assay(se, "counts")[1:5, ]
SummarizedExperiment::colData(se)
```

## Differential tRNA abundance

We can test for differential tRNA abundance between infected and uninfected
conditions using the DESeq2 wrappers.

```{r}
#| label: deseq-abundance
counts <- SummarizedExperiment::assay(se, "counts")
coldata <- as.data.frame(SummarizedExperiment::colData(se))

dds <- run_deseq(counts, coldata, design = ~condition)
res <- tidy_deseq_results(dds, contrast = c("condition", "inf", "ctl"))
```

```{r}
#| label: fig-volcano
#| fig.cap: "Volcano plot of differential tRNA abundance (inf vs ctl)."
#| fig.height: 5
plot_volcano(res)
```

### Top changing tRNAs

```{r}
#| label: top-trnas
res |>
  filter(!is.na(padj)) |>
  arrange(pvalue) |>
  head(10) |>
  select(tRNA, log2FoldChange, pvalue, padj, significant)
```

## Differential charging analysis

A unique feature of nanopore tRNA-seq is the ability to measure charging
levels. We can test whether the ratio of charged to uncharged reads changes
upon infection.

```{r}
#| label: fig-charging-ratio
#| fig.cap: "Change in charging ratio (infected - control) per tRNA."
#| fig.height: 5
charging <- S4Vectors::metadata(se)$charging
charging$condition <- ifelse(grepl("ctl", charging$sample_id), "ctl", "inf")

ratio_diff <- compute_charging_diffs(
  charging,
  numerator = "inf",
  denominator = "ctl",
  min_count = 50,
  n_top = 20
)

plot_charging_diffs(ratio_diff)
```

## Base-calling error profiles

Base-calling error rates reflect RNA modifications that cause the nanopore
basecaller to misidentify bases. Comparing error profiles between conditions
can reveal modification changes.

```{r}
#| label: bcerror-data
bcerror <- S4Vectors::metadata(se)$bcerror

# Focus on charged tRNAs (not uncharged variants)
bcerror_charged <- bcerror |>
  filter(!grepl("-uncharged$", ref))

# Calculate mean error rate per position per condition
bcerror_charged <- bcerror_charged |>
  mutate(condition = ifelse(grepl("ctl", sample_id), "ctl", "inf"))

bcerror_summary <- bcerror_charged |>
  group_by(ref, pos, condition) |>
  summarise(
    mean_error = mean(error_rate),
    mean_mis = mean(mis),
    .groups = "drop"
  )
```

```{r}
#| label: fig-bcerror-profile
#| fig.cap: "Base-calling error profiles for selected tRNAs."
#| fig.height: 7
# Plot error profiles for a few tRNAs
plot_trnas <- c(
  "host-tRNA-Ala-GGC-1-1",
  "host-tRNA-Asp-GTC-1-1",
  "host-tRNA-Glu-TTC-1-1"
)

plot_bcerror_profile(bcerror_summary, refs = plot_trnas)
```

### Error rate difference heatmap

```{r}
#| label: fig-bcerror-delta
#| fig.cap: "Difference in mean base-calling error (inf - ctl) across all tRNAs and positions."
#| fig.height: 5
# Compute delta (inf - ctl)
bcerror_delta <- bcerror_summary |>
  select(ref, pos, condition, mean_error) |>
  pivot_wider(names_from = condition, values_from = mean_error) |>
  mutate(delta = inf - ctl)

# Add Sprinzl coordinates for position ordering
sprinzl <- read_sprinzl_coords(
  clover_example("sprinzl/ecoliK12_global_coords.tsv.gz")
)

# Match tRNA names: strip "host-" prefix
bcerror_delta <- bcerror_delta |>
  mutate(trna_id = sub("^host-", "", ref))

# Join with sprinzl coords to get structural labels
bcerror_sprinzl <- bcerror_delta |>
  left_join(
    sprinzl |> select(trna_id, seq_index, sprinzl_label),
    by = c("trna_id", "pos" = "seq_index")
  ) |>
  filter(!is.na(sprinzl_label))

plot_mod_heatmap(
  bcerror_sprinzl,
  value_col = "delta",
  ref_col = "ref",
  color_limits = c(-0.1, 0.1)
)
```

## Modification annotations from MODOMICS

The [MODOMICS](https://genesilico.pl/modomics/) database catalogs known RNA
modifications. `fetch_modomics_mods()` downloads tRNA modification data for a
given organism and maps the modification positions onto your reference
sequences using pairwise alignment.

```{r}
#| label: modomics-fetch
#| message: true
fasta_path <- S4Vectors::metadata(se)$config$fasta
mods <- fetch_modomics_mods(fasta_path, organism = "Escherichia coli")
mods
```

We can overlay known modifications onto the base-calling error profiles.
Positions with known modifications often correspond to elevated error rates,
since the basecaller misidentifies modified nucleotides.

```{r}
#| label: fig-bcerror-mods
#| fig.cap: "Base-calling error profiles with known modification positions marked."
#| fig.height: 7
plot_bcerror_profile(bcerror_summary, refs = plot_trnas, mods = mods)
```

## Modification co-occurrence (odds ratios)

Odds ratios measure whether modifications at pairs of positions tend to
co-occur on the same read. Positive log odds ratios indicate co-occurrence;
negative indicate mutual exclusivity.

```{r}
#| label: odds-ratios-data
or_data <- S4Vectors::metadata(se)$odds_ratios
glimpse(or_data)
```

### Chord diagram: single sample

The Sprinzl coordinate files use RNA anticodon notation (e.g., `UUC`) while
the pipeline uses DNA notation (e.g., `TTC`). The pipeline also adds a
`host-` prefix. We need to account for these naming differences when matching.

```{r}
#| label: fig-chord-single
#| fig.cap: "Modification co-occurrence network for a single tRNA in one sample."
#| fig.height: 6
#| fig.width: 6
#| error: false
#| warning: false
# Pick one tRNA and one sample
or_single <- or_data |>
  filter(
    ref == "host-tRNA-Glu-TTC-1-1",
    sample_id == "wt-15-ctl-01"
  )

# Get sprinzl coords: note UUC (RNA) vs TTC (DNA) in anticodon
sprinzl_glu <- sprinzl |>
  filter(trna_id == "tRNA-Glu-UUC-1-1")

# Filter modifications for this tRNA
mods_glu <- mods |>
  filter(ref == "host-tRNA-Glu-TTC-1-1")

plot_chord_or(
  or_single,
  or_cutoff = 1.0,
  p_cutoff = 0.01,
  min_obs = 100,
  sprinzl_coords = sprinzl_glu,
  mods = mods_glu,
  title = "host-tRNA-Glu-TTC-1-1 (ctl-01)"
)
```

### Chord diagram: modification rewiring between conditions

We can compare modification networks between conditions using the ratio of
odds ratios (ROR). Positive log ROR indicates gained modification dependencies
in the infected condition; negative indicates lost dependencies.

```{r}
#| label: fig-chord-ror
#| fig.cap: "Modification rewiring between control and infected conditions."
#| fig.height: 6
#| fig.width: 6
#| error: false
#| warning: false
# Add condition labels
or_with_cond <- or_data |>
  mutate(condition = ifelse(grepl("ctl", sample_id), "ctl", "inf")) |>
  filter(ref == "host-tRNA-Glu-TTC-1-1")

# Compute ratio of odds ratios
ror <- compute_ror(
  or_with_cond,
  numerator = "inf",
  denominator = "ctl",
  min_obs = 100
)

plot_chord_ror(
  ror,
  ror_cutoff = 0.5,
  sprinzl_coords = sprinzl_glu,
  mods = mods_glu,
  title = "Modification rewiring: inf vs ctl (Glu-TTC-1-1)"
)
```

## Session info

```{r}
#| label: session-info
sessionInfo()
```
