---
title: "Introduction to clover"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to clover}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  warning = FALSE
)
```

```{r}
#| label: setup
#| message: false
library(clover)
library(dplyr)
library(tidyr)
```

## Overview

This vignette demonstrates clover's analysis workflow using *E. coli* tRNA
sequencing data from a T4 phage infection time-course experiment. The test
dataset includes 6 samples: 3 uninfected controls (`ctl`) and 3 T4-infected
(`inf`) wild-type *E. coli* at 15 minutes post-infection, with 3 biological
replicates per condition.

## Loading data with `create_clover()`

The simplest way to load results from the
[aa-tRNA-seq-pipeline](https://github.com/rnabioco/aa-tRNA-seq-pipeline)
is with `create_clover()`, which reads a pipeline configuration file and
assembles a `SummarizedExperiment`.

```{r}
#| label: create-se
config_path <- clover_example("ecoli/config.yaml")

sample_info <- data.frame(
  sample_id = c(
    "wt-15-ctl-01",
    "wt-15-ctl-02",
    "wt-15-ctl-03",
    "wt-15-inf-01",
    "wt-15-inf-02",
    "wt-15-inf-03"
  ),
  condition = rep(c("ctl", "inf"), each = 3),
  replicate = rep(1:3, 2)
)

se <- create_clover(
  config_path,
  types = c("charging", "bcerror", "odds_ratios"),
  sample_info = sample_info
)

se
```

The SE object contains:

- **assay "counts"**: abundance count matrix (charged + uncharged reads)
- **colData**: sample metadata with condition labels
- **metadata**: list with raw charging data, bcerror, odds ratios, FASTA reference, and pipeline config

```{r}
#| label: inspect-se
SummarizedExperiment::assay(se, "counts")[1:5, ]
SummarizedExperiment::colData(se)
```

## Differential tRNA abundance

We can test for differential tRNA abundance between infected and uninfected
conditions using the DESeq2 wrappers.

```{r}
#| label: deseq-abundance
counts <- SummarizedExperiment::assay(se, "counts")
coldata <- as.data.frame(SummarizedExperiment::colData(se))

dds <- run_deseq(counts, coldata, design = ~condition)
res <- tidy_deseq_results(dds, contrast = c("condition", "inf", "ctl"))
```

```{r}
#| label: fig-volcano
#| fig.cap: "Volcano plot of differential tRNA abundance (inf vs ctl)."
#| fig.height: 5
plot_volcano(res)
```

### Top changing tRNAs

```{r}
#| label: top-trnas
tabulate_deseq(res)
```

## Differential charging analysis

A unique feature of nanopore tRNA-seq is the ability to measure charging
levels. We can test whether the ratio of charged to uncharged reads changes
upon infection.

```{r}
#| label: fig-charging-ratio
#| fig.cap: "Change in charging ratio (infected - control) per tRNA."
#| fig.height: 5
charging <- S4Vectors::metadata(se)$charging
charging$condition <- ifelse(grepl("ctl", charging$sample_id), "ctl", "inf")

ratio_diff <- compute_charging_diffs(
  charging,
  numerator = "inf",
  denominator = "ctl",
  min_count = 50,
  n_top = 20
)

plot_charging_diffs(ratio_diff)
```

### Abundance versus charging

We can visualize the relationship between abundance changes and charging
ratio changes on a single scatter plot. This highlights tRNAs where
expression and aminoacylation are coordinately or discordantly affected.

```{r}
#| label: fig-abundance-charging
#| fig.cap: "Abundance change versus charging ratio change per tRNA."
#| fig.height: 6
plot_abundance_charging(res, ratio_diff)
```

## Base-calling error profiles

Base-calling error rates reflect RNA modifications that cause the nanopore
basecaller to misidentify bases. Comparing error profiles between conditions
can reveal modification changes.

```{r}
#| label: bcerror-data
# Load preprocessed bcerror data comparing wt vs TruB-del (tb) for 10 tRNAs
bcerror_rds <- readRDS(clover_example("ecoli/bcerror_summary.rds"))
bcerror_charged <- bcerror_rds$bcerror_charged
bcerror_summary <- bcerror_rds$bcerror_summary
```

```{r}
#| label: fig-bcerror-profile
#| fig.cap: "Base-calling error profiles for selected tRNAs."
#| fig.height: 7
# Plot error profiles for a few tRNAs
plot_trnas <- c(
  "host-tRNA-Asn-GTT-1-1",
  "host-tRNA-Glu-TTC-1-1",
  "host-tRNA-Phe-GAA-1-1"
)

plot_bcerror_profile(bcerror_summary, refs = plot_trnas)
```

### Error rate difference heatmap

```{r}
#| label: fig-bcerror-delta
#| fig.cap: "Difference in mean base-calling error (wt - TruB-del) across all tRNAs and positions."
#| fig.height: 5
#| message: false
# Compute delta (wt - tb); positive = higher error in wt (modification present)
bcerror_delta <- compute_bcerror_delta(bcerror_summary, delta = wt - tb)

# Add Sprinzl coordinates and MODOMICS modification annotations
sprinzl <- read_sprinzl_coords(
  clover_example("sprinzl/ecoliK12_global_coords.tsv.gz")
)
trna_fasta <- clover_example("ecoli/trna_only.fa.gz")
mods <- modomics_mods(trna_fasta, organism = "Escherichia coli")

heatmap_data <- prep_mod_heatmap(
  bcerror_delta,
  sprinzl_coords = sprinzl,
  mods = mods
)

plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label",
  cluster = FALSE,
  color_limits = c(-0.1, 0.1),
  highlight_col = "has_mod"
)
```

## Modification annotations from MODOMICS

The [MODOMICS](https://genesilico.pl/modomics/) database catalogs known RNA
modifications. We already used `modomics_mods()` above to annotate the
heatmap with known modification positions. This function maps MODOMICS
entries onto your reference sequences using pairwise alignment. Data for
common organisms is bundled with the package (see `modomics_organisms()`),
so no internet connection is needed.

```{r}
#| label: modomics-fetch
mods
```

We can overlay known modifications onto the base-calling error profiles.
Positions with known modifications often correspond to elevated error rates,
since the basecaller misidentifies modified nucleotides.

```{r}
#| label: fig-bcerror-mods
#| fig.cap: "Base-calling error profiles with known modification positions marked."
#| fig.height: 7
plot_bcerror_profile(bcerror_summary, refs = plot_trnas, mods = mods)
```

## tRNA secondary structure visualization

clover includes pre-computed tRNA cloverleaf SVGs for several model organisms.
You can overlay modification annotations, circle outlines, and text color
changes onto these structures.

```{r}
#| label: structure-organisms
# List available organisms and tRNAs
structure_organisms()
head(structure_trnas("Escherichia coli"))
```

### Base structure

The simplest call renders the bare cloverleaf with base-pair lines and
nucleotide letters.

```{r}
#| label: structure-bare
svg_path <- plot_tRNA_structure("tRNA-Glu-TTC", "Escherichia coli")
```

```{r}
#| label: fig-structure-bare
#| echo: false
structure_html(svg_path)
```

### Annotated structure with modifications, outlines, and text colors

We can highlight the anticodon with filled circles, mark the discriminator
base with an outline, and change the text color of specific positions.

```{r}
#| label: structure-annotated
# Known modifications for this tRNA
mods_glu_struct <- mods |>
  filter(ref == "host-tRNA-Glu-TTC-1-1")

# Anticodon positions (34-36) as filled circles
anticodon <- tibble::tibble(
  pos = 34:36,
  mod1 = rep("anticodon", 3)
)

# Combine MODOMICS mods with anticodon highlight
all_mods <- bind_rows(mods_glu_struct, anticodon)

# Discriminator base (position 73) as an outline
disc_outline <- tibble::tibble(pos = 73, group = "discriminator")

# Change discriminator text to red
disc_text <- tibble::tibble(pos = 73, color = "#E41A1C")

svg_path <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  modifications = all_mods,
  mod_palette = c(default_mod_palette(), anticodon = "#4DAF4A"),
  outlines = disc_outline,
  outline_palette = c(discriminator = "#E41A1C"),
  text_colors = disc_text
)
```

```{r}
#| label: fig-structure-annotated
#| echo: false
structure_html(svg_path)
```

### Long variable arm tRNAs

Leu and Ser tRNAs have a 4-stem cloverleaf with a variable arm stem,
which is properly rendered from the tRNAscan-SE covariance model alignment.

```{r}
#| label: structure-leu
svg_path <- plot_tRNA_structure("tRNA-Leu-CAA", "Escherichia coli")
```

```{r}
#| label: fig-structure-leu
#| echo: false
structure_html(svg_path)
```

### Modification co-occurrence on structure

We can visualize pairwise modification co-occurrence directly on the tRNA
cloverleaf. Arcs connect position pairs with significant odds ratios:
blue arcs indicate mutual exclusivity (negative log OR) and vermillion
arcs indicate co-occurrence (positive log OR). Stroke width encodes the
magnitude.

```{r}
#| label: structure-linkages
or_data <- S4Vectors::metadata(se)$odds_ratios

# Filter OR data for one tRNA and one sample, clean, and filter for
# significance. filter_linkages() returns a tibble ready for plotting.
linkages_glu <- or_data |>
  filter(
    ref == "host-tRNA-Glu-TTC-1-1",
    sample_id == "wt-15-ctl-01"
  ) |>
  clean_odds_ratios() |>
  filter_linkages()

svg_path <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  modifications = mods_glu_struct,
  linkages = linkages_glu
)
```

```{r}
#| label: fig-structure-linkages
#| echo: false
structure_html(svg_path)
```

## Next steps

For isodecoder-level modification rewiring analysis — including odds ratio
aggregation, ratio of odds ratios, dimensionality reduction, and network
visualization — see `vignette("rewiring", package = "clover")`.

## Session info

```{r}
#| label: session-info
sessionInfo()
```
