----
title: build a s4 class data structure
description: generate matricies that will be used to combine into a standard summarized experiment object
----

# build a s4 data frame using bioconductor in R
# create this data frame similar to a summarized experiment

# test data: wildtype charged ecoli sequencing 30 minutes uninfected vs infected (MOI = 10)

```{r}
load_pkgs <- function(pkgs){
    failed_pkgs <- c() # Initialize a vector to store failed installations
    for (pkg in pkgs) {
        library(pkg, character.only = TRUE, quietly = TRUE)

    }
    failed_pkgs <<- c(failed_pkgs, pkg)
        message(sprintf("Failed to load package: %s", pkg))

    if (length(failed_pkgs) > 0) {
        message("The following packages failed to install or load:")
        print(failed_pkgs)
    } else {
        message("All packages installed and loaded successfully.")
    }  
}
pkgs <- c("tidyverse",
          "here",
          "cowplot",
          "purrr",
          "ggrepel",
          "patchwork",
          "FNN",
          "fs",
          "glue",
          "janitor",
          "ggokabeito", 
          "fs",
          "SummarizedExperiment")

load_pkgs(pkgs)            
```

# summarized experiment
```{r}
library(SummarizedExperiment)

```



# Load in data in a simple format
# TODO:
1. remove unneccessary data files from ecoli folder (done)
2. generate a readme.md to keep track of the files (done)
3. set seed and use a small subset of data to test
4. remove dates from sample names (done)
 # the idea is that the function will take a tsv file and then generate the summarized experiment. Under the hood i will have to mannully create the matricies that will be joined in to create the object 'metadata', 'assays' etc. 

```{r}
# build file names
samples <- c("inf", "uninf")
replicates <- "r1"
counts_files <- crossing (id = samples, repl = replicates) |> 
    mutate(path = path(here(), 'inst/extdata/ecoli', glue("{id}-{repl}.charging.cpm.tsv.gz")))

bc_files <- crossing (id = sample_ids, repl = replicates) |> 
    mutate(path = path(here(), 'inst/extdata/ecoli', glue("{id}-{repl}.bcerror.tsv.gz")))


```

 # function below is not generalizable but should work
```{r}
library(SummarizedExperiment)
# 1. First need to bring in data files and clean them.
create_df_from_files <- function(results_dir, extension){
  """ 
  this function can handle cpm counts and basecalling error bcerror.
  it takes a results directory and file extension to access and read each file that matches the given description.
  it will then conditionally clean the data frame depending on which type of extension it is.
  I am not sure this will be generalizable in the future, however, the data frame does need to be built and cleaned prior to S4 construction
  """


  # file pattern
  pattern <- glue("\\.{extension}\\.tsv\\.gz$")

  # Get all file paths recursively
  files <- list.files(results_dir, pattern = pattern, full.names = TRUE, recursive = TRUE)

  # create dataframe 
  df <- map_dfr(files, ~ read_tsv(gzfile(.x)) |> 
                 mutate(sample = basename(.x)) |> 
                 mutate(sample = str_remove(sample, glue("\\.{extension}\\.tsv\\.gz$")))) # Strip file suffix
  
  # clean column names
  df <- clean_names(df)

  # wrangle dataframe:
  # clean reference name
  # clean sample name
  # summarize tRNA counts into family - could reduce granularity
  # add MOI column 
  if (extension == "charging.cpm"){
    df <- df |> 
    separate_wider_delim(sample, 
                        delim = "-", 
                        names = c('date', 'infection_status', 'genotype', 'timepoint')) |> 
    separate_wider_delim(t_rna, 
                        delim = "-", 
                        names = c("origin", "trna", "aa", "anticodon", "family", "species"), 
                        too_many = "drop",
                        too_few = "align_start") |> # exclude uncharged ref for now
    select(-trna) |>
    group_by(origin, aa, anticodon, species, family, genotype, infection_status, timepoint, date) |> 
    summarize(
    charged = sum(charged),
    uncharged = sum(uncharged),
    charged_cpm = sum(charged_cpm),
    uncharged_cpm = sum(uncharged_cpm),
    .groups = 'drop'  # Optionally ungroup after summarizing
    ) |> 
    # specific to ecoli data
    mutate(moi = case_when(
      date == "20240805" ~ "one",
      date == "20240813" ~ "one",
      date == "20241018" ~ "one",
      date == "20241219" ~ "ten"
    )) |> 
    mutate(isodecoder = paste(origin, aa, anticodon, sep = "-"),
    reference = paste(origin, aa, anticodon, species, family, sep = "-"))

    charged_cpm <- df |> 
        filter(infection_status == "uninf") |> 
        select(genotype, reference, charged_cpm) |> 
        pivot_wider(
            names_from = genotype, 
            values_from = charged_cpm,
            values_fill = 0
        ) |> 
        column_to_rownames("reference") |> 
        as.matrix()

    uncharged_cpm <- df |> 
        filter(infection_status == "uninf") |> 
        select(genotype, reference, uncharged_cpm) |> 
        pivot_wider(
            names_from = genotype, 
            values_from = uncharged_cpm,
            values_fill = 0
        ) |> 
        column_to_rownames("reference") |> 
        as.matrix()
  } 


  # wrangle dataframe:
  # clean reference names
  # calculate bcerror
  # clean sample column - make date, moi, sample columns
  # create isodecoder column 
  # correct position for 24 nt 3' adapter
  if (extension == "bcerror"){
    df <- df |> 
      separate_wider_delim(sample, 
                        delim = "-", 
                        names = c('date', 'infection_status', 'genotype', 'timepoint')) |> 
      separate_wider_delim(reference, 
                        delim = "-", 
                        names = c("origin", "trna", "aa", "anticodon", "family", "species"), 
                        too_many = "drop",
                        too_few = "align_start") |>
      select(-trna) |> 
      mutate(position = position -24, 
            isodecoder = paste(origin, aa, anticodon, sep = "-"))  
  }
  


  # return data frame
  return(df, charged_cpm, uncharged_cpm)
}

results_dir <- here("inst/extdata/ecoli")
cpm_extension <- "charging.cpm"
bc_extension <- "bcerror"

cpm_df <- create_df_from_files(results_dir, cpm_extension)
bcerr_df < create_df_from_files(results_dir, bc_extension)

# step 2:
# create matricies for each read category: charged charged-cpm uncharged uncharged-cpm infected and uninfected
# reshape data for raw and normalized read counts
matrix_maker <- function(df, filter, status){
    df <- df |> 
        filter(infection_status == filter) |> 
        select(genotype, isodecoder, {{status}}) |> 
        group_by(genotype, isodecoder) |> 
        summarize(
            charged = sum({{status}}),
            .groups = 'drop'
        ) |> 
        pivot_wider(
            names_from = genotype, 
            values_from = {{status}},
            values_fill = 0
        ) |> 
        column_to_rownames("isodecoder") |> 
        as.matrix()

    return(df)
}
test <- matrix_maker(cpm_df, "uninf", "charged")


```