---
title: "Plotting modification heatmaps"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  warning = FALSE
)
```

```{r}
#| label: setup
#| message: false
library(clover)
library(dplyr)
```

## Overview

This article is a visual gallery for clover's modification heatmap functions.
It covers every parameter of `plot_mod_heatmap()` and `prep_mod_heatmap()`,
plus the related `plot_bcerror_profile()` and `plot_mod_landscape()` functions.

All examples use the bundled *E. coli* base-calling error data shipped with
the package.

## Data preparation

The bundled RDS contains pre-summarized bcerror data comparing wild-type
(`wt`) and TruB-deletion (`tb`) *E. coli* across 10 tRNAs.

```{r}
#| label: load-data
bcerror_rds <- readRDS(clover_example("ecoli/bcerror_summary.rds"))
bcerror_summary <- bcerror_rds$bcerror_summary
bcerror_charged <- bcerror_rds$bcerror_charged
```

Compute the per-position delta (wt minus TruB-del). Positive values mean
higher error in wild-type, indicating a modification lost in the mutant.

```{r}
#| label: compute-delta
bcerror_delta <- compute_bcerror_delta(bcerror_summary, delta = wt - tb)
bcerror_delta
```

## Preparing heatmap data

`prep_mod_heatmap()` joins Sprinzl coordinates, optionally annotates known
modifications from MODOMICS, and shortens tRNA names for display.

```{r}
#| label: prep-heatmap
#| message: false
sprinzl <- read_sprinzl_coords(
  clover_example("sprinzl/ecoliK12_global_coords.tsv.gz")
)
trna_fasta <- clover_example("ecoli/trna_only.fa.gz")
mods <- modomics_mods(trna_fasta, organism = "Escherichia coli")

heatmap_data <- prep_mod_heatmap(
  bcerror_delta,
  sprinzl_coords = sprinzl,
  mods = mods
)

heatmap_data
```

The result is a tibble with `sprinzl_label` as an ordered factor (canonical
tRNA position ordering), `trna_label` for compact y-axis names, and `has_mod`
marking known MODOMICS modification sites.

## Basic heatmap

The simplest call passes the prepared data with appropriate column mappings.

```{r}
#| label: fig-basic
#| fig.cap: "Basic modification heatmap with default settings."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label"
)
```

## Color scale

Control the diverging color scale with `color_limits`, `color_low`,
`color_high`, and `na_value`.

```{r}
#| label: fig-color-scale
#| fig.cap: "Custom color scale with tighter limits and purple-green palette."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1),
  color_low = "#7B3294",
  color_high = "#008837",
  na_value = "gray95"
)
```

The default palette uses colorblind-friendly blue (`#0072B2`) and vermillion
(`#D55E00`). Values outside `color_limits` are squished to the extremes.

## Clustering

By default, rows are clustered using Ward's D2 hierarchical clustering on
the Sprinzl-position value matrix. Set `cluster = FALSE` to keep alphabetical
order.

```{r}
#| label: fig-cluster-on
#| fig.cap: "Clustered rows (default)."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1)
)
```

```{r}
#| label: fig-cluster-off
#| fig.cap: "Alphabetical row order (cluster = FALSE)."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label",
  cluster = FALSE,
  color_limits = c(-0.1, 0.1)
)
```

### Noise filtering with `cluster_threshold`

When clustering, low-level noise can dominate the distance matrix. Setting
`cluster_threshold` restricts clustering to positions where at least one
tRNA exceeds the threshold in absolute value.

```{r}
#| label: fig-cluster-threshold
#| fig.cap: "Clustering uses only positions with |delta| > 0.05."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1),
  cluster_threshold = 0.05
)
```

## Group-aware clustering

When tRNAs belong to natural groups (e.g., amino acid families), use
`group_col` to cluster within each group and draw horizontal dividers
between them. First, add an amino acid column to the data.

```{r}
#| label: add-amino-acid
heatmap_grouped <- heatmap_data |>
  mutate(amino_acid = sub("-.*", "", trna_label))
```

```{r}
#| label: fig-group-cluster
#| fig.cap: "Clustering within amino acid groups, separated by dividers."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_grouped,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1),
  group_col = "amino_acid"
)
```

Control the divider line thickness with `divider_linewidth`.

```{r}
#| label: fig-divider-width
#| fig.cap: "Thicker dividers between groups."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_grouped,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1),
  group_col = "amino_acid",
  divider_linewidth = 2
)
```

## Modification highlights

When `prep_mod_heatmap()` is called with `mods`, a logical `has_mod` column
marks known modification sites. Use `highlight_col` to overlay dots on those
cells.

```{r}
#| label: fig-highlight
#| fig.cap: "MODOMICS modification sites marked with dots."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1),
  highlight_col = "has_mod"
)
```

Adjust dot appearance with `highlight_size` and `highlight_offset` (x/y
displacement from tile center).

```{r}
#| label: fig-highlight-custom
#| fig.cap: "Larger highlight dots with custom offset."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1),
  highlight_col = "has_mod",
  highlight_size = 1.2,
  highlight_offset = c(-0.3, 0.3)
)
```

## Tile labels

Overlay text on tiles with `label_col`. Labels are only shown where
`abs(value) >= label_min`, and text color automatically switches between
black and white for contrast.

```{r}
#| label: add-ref-base
#| message: false
# Add the reference base at each position for labeling
ref_seqs <- Biostrings::readDNAStringSet(trna_fasta)
ref_bases <- tibble::tibble(
  ref = rep(names(ref_seqs), Biostrings::width(ref_seqs)),
  pos = unlist(lapply(Biostrings::width(ref_seqs), seq_len)),
  ref_base = strsplit(as.character(unlist(ref_seqs)), "") |> unlist()
)

heatmap_labeled <- heatmap_data |>
  left_join(ref_bases, by = c("ref", "pos"))
```

```{r}
#| label: fig-labels
#| fig.cap: "Reference bases overlaid on tiles where |delta| >= 0.03."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_labeled,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1),
  label_col = "ref_base",
  label_min = 0.03,
  label_size = 2
)
```

## Legend and caption

Customize the fill legend with `fill_name` and `fill_breaks`, and add
explanatory text with `caption`.

```{r}
#| label: fig-legend-caption
#| fig.cap: "Custom legend title, breaks, and caption."
#| fig.height: 5
plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1),
  fill_name = "\u0394 error rate",
  fill_breaks = c(-0.1, -0.05, 0, 0.05, 0.1),
  caption = "Positive values = higher error in wt (modification present in wt, lost in TruB-del)."
)
```

## Square versus free aspect ratio

By default, `square = TRUE` uses `coord_fixed(ratio = 1)` for square tiles.
Set `square = FALSE` to let ggplot2 fill the available space, which is useful
for wide heatmaps with many positions.

```{r}
#| label: fig-square-false
#| fig.cap: "Free aspect ratio stretches tiles to fill the plot area."
#| fig.height: 4
plot_mod_heatmap(
  heatmap_data,
  value_col = "delta",
  ref_col = "trna_label",
  color_limits = c(-0.1, 0.1),
  square = FALSE
)
```

## Base-calling error profiles

`plot_bcerror_profile()` creates line plots of per-position error rates,
faceted by tRNA and colored by condition.

```{r}
#| label: fig-profile-basic
#| fig.cap: "Error profiles for three tRNAs."
#| fig.height: 7
plot_trnas <- c(
  "host-tRNA-Asn-GTT-1-1",
  "host-tRNA-Glu-TTC-1-1",
  "host-tRNA-Phe-GAA-1-1"
)

plot_bcerror_profile(bcerror_summary, refs = plot_trnas)
```

### Adding modification annotations

Pass a `mods` tibble to overlay vertical dashed lines at known modification
positions.

```{r}
#| label: fig-profile-mods
#| fig.cap: "Error profiles with MODOMICS modification positions marked."
#| fig.height: 7
plot_bcerror_profile(bcerror_summary, refs = plot_trnas, mods = mods)
```

### Custom colors and layout

Override condition colors with a named vector, and change the facet layout
with `ncol`.
```{r}
#| label: fig-profile-custom
#| fig.cap: "Custom colors and two-column layout."
#| fig.height: 5
plot_bcerror_profile(
  bcerror_summary,
  refs = plot_trnas,
  mods = mods,
  colors = c(wt = "#E69F00", tb = "#56B4E9"),
  ncol = 2
)
```

## Modification landscape

`plot_mod_landscape()` creates a stacked panel plot showing multiple metrics
along the tRNA sequence. Each metric gets its own panel sharing a common
x-axis. This is useful for comparing error rates, mismatch types, or other
per-position statistics side by side.

First, prepare single-tRNA data with Sprinzl coordinates and region
annotations.

```{r}
#| label: prep-landscape
#| message: false
# Pick one tRNA and reshape for landscape plotting
landscape_data <- bcerror_summary |>
  filter(ref == "host-tRNA-Glu-TTC-1-1", condition == "wt")

# Add Sprinzl coordinates and region info
sprinzl_glu <- sprinzl |>
  filter(trna_id == "tRNA-Glu-UUC-1-1") |>
  select(pos, sprinzl_label, region)

landscape_data <- landscape_data |>
  left_join(sprinzl_glu, by = "pos") |>
  filter(!is.na(sprinzl_label))

# Add a logical column for known modification positions
mod_positions <- mods |>
  filter(ref == "host-tRNA-Glu-TTC-1-1") |>
  distinct(pos) |>
  mutate(has_mod = TRUE)

landscape_data <- landscape_data |>
  left_join(mod_positions, by = "pos") |>
  mutate(has_mod = tidyr::replace_na(has_mod, FALSE))
```

```{r}
#| label: fig-landscape-basic
#| fig.cap: "Stacked landscape of error rate and mismatch frequency."
#| fig.height: 5
plot_mod_landscape(
  landscape_data,
  metrics = c("mean_error", "mean_mis")
)
```

### Region shading

Pass `region_col` to add background shading for structural regions
(acceptor-stem, D-loop, anticodon-stem, etc.).

```{r}
#| label: fig-landscape-regions
#| fig.cap: "Landscape with structural region shading."
#| fig.height: 5
plot_mod_landscape(
  landscape_data,
  metrics = c("mean_error", "mean_mis"),
  region_col = "region"
)
```

### Sprinzl axis and modification highlights

Add a secondary x-axis with Sprinzl position labels using `sprinzl_col`.
Use `mod_col` to highlight known modification positions on that axis (shown
in bold orange when ggtext is installed).

```{r}
#| label: fig-landscape-full
#| fig.cap: "Full landscape with regions, Sprinzl axis, and modification highlights."
#| fig.height: 5
plot_mod_landscape(
  landscape_data,
  metrics = c("mean_error", "mean_mis"),
  region_col = "region",
  sprinzl_col = "sprinzl_label",
  mod_col = "has_mod",
  title = "tRNA-Glu-UUC-1-1 modification landscape",
  heights = c(2, 1)
)
```

## Session info

```{r}
#| label: session-info
sessionInfo()
```
