---
title: "Modification rewiring analysis"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  warning = FALSE
)
```

```{r}
#| label: setup
#| message: false
library(clover)
library(dplyr)
```

## Overview

This article demonstrates clover's modification rewiring analysis workflow.
"Rewiring" refers to changes in modification co-occurrence patterns between
conditions — some position pairs may gain or lose coordinated modification.

The workflow operates at the **isodecoder** level, collapsing gene copies
to increase statistical power. It uses the same *E. coli* T4 phage infection
dataset as the main vignette.

## Load data

```{r}
#| label: load-data
config_path <- clover_example("ecoli/config.yaml")

sample_info <- data.frame(
  sample_id = c(
    "wt-15-ctl-01",
    "wt-15-ctl-02",
    "wt-15-ctl-03",
    "wt-15-inf-01",
    "wt-15-inf-02",
    "wt-15-inf-03"
  ),
  condition = rep(c("ctl", "inf"), each = 3)
)

se <- create_clover(
  config_path,
  types = c("charging", "odds_ratios"),
  sample_info = sample_info
)

or_data <- S4Vectors::metadata(se)$odds_ratios
or_data$condition <- ifelse(grepl("ctl", or_data$sample_id), "ctl", "inf")
```

## Clean and aggregate odds ratios

The raw odds ratios may contain infinite values from zero-cell contingency
tables. `clean_odds_ratios()` caps these at a reasonable threshold.

```{r}
#| label: clean-or
or_clean <- clean_odds_ratios(or_data)
glimpse(or_clean)
```

Next, collapse per-gene odds ratios to the isodecoder level. This averages
across gene copies (e.g., `tRNA-Glu-TTC-1-1` and `tRNA-Glu-TTC-2-1` become
`tRNA-Glu-TTC`).

```{r}
#| label: aggregate-isodecoder
or_ctl <- or_clean |>
  filter(condition == "ctl") |>
  aggregate_or_isodecoder()

or_inf <- or_clean |>
  filter(condition == "inf") |>
  aggregate_or_isodecoder()

or_ctl
```

## Compare conditions with relative odds ratios

`compute_ror_isodecoder()` computes the relative odds ratio (ROR) between
two conditions with z-score significance testing.

```{r}
#| label: compute-ror
ror <- compute_ror_isodecoder(or_inf, or_ctl)
ror
```

## Rewiring scores and dimensionality reduction

Build a matrix of ROR values (isodecoders × position pairs) and compute
per-isodecoder rewiring scores.

```{r}
#| label: rewiring-matrix
mat <- prepare_rewiring_matrix(ror)
scores <- calculate_rewiring_scores(mat)
scores |> arrange(desc(euclidean_magnitude))
```

Run PCoA for dimensionality reduction and visualize the result.

```{r}
#| label: fig-pcoa
#| fig.cap: "PCoA of tRNA modification rewiring between control and infected conditions."
#| fig.height: 6
pcoa <- perform_pcoa(mat)
plot_pcoa_rewiring(pcoa, scores)
```

## Network visualization

Build a modification network from the ROR data and visualize it as an arc
diagram.

```{r}
#| label: fig-arc-diagram
#| fig.cap: "Arc diagram of modification rewiring network."
#| fig.height: 6
graph <- build_or_network(ror)
plot_arc_diagram(graph, title = "Modification rewiring: inf vs ctl")
```

## Session info

```{r}
#| label: session-info
sessionInfo()
```
