---
title: "Plotting tRNA structures"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  warning = FALSE
)
```

```{r}
#| label: setup
#| message: false
library(clover)
library(dplyr)
```

## Overview

clover includes pre-computed tRNA cloverleaf SVGs that can be annotated with
modification highlights, outline circles, text color changes, and linkage arcs.
This vignette walks through every customization option for
`plot_tRNA_structure()` and related functions.

## Discovering available structures

Bundled SVGs are organized by organism. Use `structure_organisms()` to list
available organisms and `structure_trnas()` to list tRNAs within an organism.

```{r}
#| label: available-organisms
structure_organisms()
```

```{r}
#| label: available-trnas
head(structure_trnas("Escherichia coli"))
```

## Basic structure

The simplest call renders the bare cloverleaf with base-pair lines,
nucleotide letters, and position markers every 10 nucleotides.

```{r}
#| label: basic-structure
svg <- plot_tRNA_structure("tRNA-Glu-TTC", "Escherichia coli")
```

```{r}
#| label: fig-basic-structure
#| echo: false
structure_html(svg)
```

## Position markers

By default, grey position numbers are drawn every 10 nucleotides. Set
`position_markers = FALSE` to hide them.

```{r}
#| label: no-markers
svg <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  position_markers = FALSE
)
```

```{r}
#| label: fig-no-markers
#| echo: false
structure_html(svg)
```

## Modification highlights

Modifications are shown as colored filled circles behind nucleotide letters.
The `modifications` argument takes a tibble with `pos` and `mod1` columns.
Output from `modomics_mods()` works directly after filtering to the tRNA of
interest.

```{r}
#| label: mods-setup
fasta_path <- clover_example("ecoli/trna_only.fa.gz")
mods <- modomics_mods(fasta_path, "Escherichia coli")

mods_glu <- mods |>
  filter(ref == "host-tRNA-Glu-TTC-1-1")
mods_glu
```

Since `plot_tRNA_structure()` uses 1-based sequence positions, we load Sprinzl
coordinates to convert named positions (like the anticodon at Sprinzl 34-36)
to their actual sequence indices.

```{r}
#| label: sprinzl-setup
sprinzl <- read_sprinzl_coords(
  clover_example("sprinzl/ecoliK12_global_coords.tsv.gz")
)
glu_coords <- sprinzl |>
  filter(trna_id == "tRNA-Glu-UUC-1-1")
```

```{r}
#| label: mods-highlight
svg <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  modifications = mods_glu
)
```

```{r}
#| label: fig-mods-highlight
#| echo: false
structure_html(svg)
```

## Custom modification palettes

`default_mod_palette()` returns a named character vector mapping modification
short names to colors. You can extend or override it.

```{r}
#| label: default-palette
head(default_mod_palette(), 10)
```

To add a custom category (e.g., highlighting anticodon positions), combine
MODOMICS modifications with your own entries and extend the palette.

```{r}
#| label: custom-palette
# Add anticodon positions (Sprinzl 34-36) as a custom category
ac_pos <- glu_coords |>
  filter(sprinzl_label %in% c("34", "35", "36")) |>
  pull(pos)

anticodon <- tibble::tibble(
  pos = ac_pos,
  mod1 = rep("anticodon", 3)
)

all_mods <- bind_rows(mods_glu, anticodon)

# Extend the palette with a color for the new category
custom_palette <- c(default_mod_palette(), anticodon = "#4DAF4A")

svg <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  modifications = all_mods,
  mod_palette = custom_palette
)
```

```{r}
#| label: fig-custom-palette
#| echo: false
structure_html(svg)
```

## Outline circles

Outlines draw stroke-only circles around nucleotides, useful for highlighting
positions without covering the base letter. The `outlines` argument takes a
tibble with `pos` and `group` columns; `outline_palette` maps group names to
colors.

```{r}
#| label: outlines
# Highlight the discriminator base (Sprinzl position 73)
disc_pos <- glu_coords |>
  filter(sprinzl_label == "73") |>
  pull(pos)

disc_outline <- tibble::tibble(pos = disc_pos, group = "discriminator")

svg <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  outlines = disc_outline,
  outline_palette = c(discriminator = "#E41A1C")
)
```

```{r}
#| label: fig-outlines
#| echo: false
structure_html(svg)
```

## Text colors

Use `text_colors` to change the color of nucleotide letters at specific
positions. This takes a tibble with `pos` and `color` columns.

```{r}
#| label: text-colors
# Color the anticodon triplet (Sprinzl 34-36) green
ac_text <- tibble::tibble(
  pos = ac_pos,
  color = rep("#4DAF4A", 3)
)

svg <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  text_colors = ac_text
)
```

```{r}
#| label: fig-text-colors
#| echo: false
structure_html(svg)
```

## Combining annotations

Modifications, outlines, and text colors can all be used together.

```{r}
#| label: combined
disc_outline <- tibble::tibble(pos = disc_pos, group = "discriminator")
disc_text <- tibble::tibble(pos = disc_pos, color = "#E41A1C")

svg <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  modifications = all_mods,
  mod_palette = custom_palette,
  outlines = disc_outline,
  outline_palette = c(discriminator = "#E41A1C"),
  text_colors = disc_text
)
```

```{r}
#| label: fig-combined
#| echo: false
structure_html(svg)
```

## Linkage arcs

Pairwise modification co-occurrence can be visualized as arcs on the
cloverleaf. The `linkages` argument takes a tibble with `pos1`, `pos2`, and
optionally `value` columns. If a `log_odds_ratio` column is present (as in
output from `clean_odds_ratios()` / `filter_linkages()`), it is used
automatically.

Arc color encodes direction: blue for mutual exclusivity (negative log OR)
and vermillion for co-occurrence (positive log OR). Stroke width encodes
magnitude.

```{r}
#| label: linkages-setup
or_path <- clover_example(
  "ecoli/summary/tables/wt-15-ctl-01/wt-15-ctl-01.odds_ratios_filtered.tsv.gz"
)
or_data <- read_odds_ratios(or_path)

linkages_glu <- or_data |>
  filter(ref == "host-tRNA-Glu-TTC-1-1") |>
  clean_odds_ratios() |>
  filter_linkages()
```

```{r}
#| label: linkages-plot
svg <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  modifications = mods_glu,
  linkages = linkages_glu
)
```

```{r}
#| label: fig-linkages-plot
#| echo: false
structure_html(svg)
```

## Custom linkage palettes

The `linkage_palette` parameter is a character vector of length 2: the first
color is used for negative values (exclusive), the second for positive values
(co-occurring). The default is `c("#0072B2", "#D55E00")`.

```{r}
#| label: custom-linkage-palette
svg <- plot_tRNA_structure(
  "tRNA-Glu-TTC",
  "Escherichia coli",
  linkages = linkages_glu,
  linkage_palette = c("#7570B3", "#D95F02")
)
```

```{r}
#| label: fig-custom-linkage-palette
#| echo: false
structure_html(svg)
```

## Long variable arm tRNAs

Leu and Ser tRNAs have a long variable arm with an extra stem-loop. These
are properly rendered from the tRNAscan-SE covariance model alignment.

```{r}
#| label: long-arm
svg <- plot_tRNA_structure("tRNA-Leu-CAA", "Escherichia coli")
```

```{r}
#| label: fig-long-arm
#| echo: false
structure_html(svg)
```

```{r}
#| label: long-arm-ser
svg <- plot_tRNA_structure("tRNA-Ser-GGA", "Escherichia coli")
```

```{r}
#| label: fig-long-arm-ser
#| echo: false
structure_html(svg)
```

## Identity elements

`plot_identity_structure()` overlays experimentally validated aminoacylation
identity elements (from Giege & Eriani, 2023) onto the cloverleaf. Strong
determinants are outlined in red, weak determinants in blue.

This requires Sprinzl coordinates to map element positions onto the tRNA
sequence. Bundled coordinate files are available for *E. coli* and
*S. cerevisiae*.

```{r}
#| label: identity-single
coords <- read_sprinzl_coords(
  clover_example("sprinzl/sacCer_global_coords.tsv.gz")
)

svg <- plot_identity_structure(
  "tRNA-Ala-AGC",
  "Saccharomyces cerevisiae",
  coords
)
```

```{r}
#| label: fig-identity-single
#| echo: false
structure_html(svg)
```

### Identity element panel

`plot_identity_panel()` arranges multiple tRNAs in a grid, each annotated
with its amino acid family's identity elements. Use `ncol` to control the
number of columns.

```{r}
#| label: identity-panel
svg <- plot_identity_panel(
  c("tRNA-Ala-AGC", "tRNA-Asp-GTC", "tRNA-Phe-GAA", "tRNA-His-GTG"),
  "Saccharomyces cerevisiae",
  coords,
  ncol = 2
)
```

```{r}
#| label: fig-identity-panel
#| echo: false
structure_html(svg)
```

## Output formats

`plot_tRNA_structure()` writes an SVG file and returns the path invisibly.
For embedding in R Markdown or Quarto, use `structure_html()` to wrap the
SVG in a centered `<div>`. For raster output, use `structure_to_png()`.

```{r}
#| label: output-formats
#| eval: false
# Embed as HTML (used throughout this vignette)
svg <- plot_tRNA_structure("tRNA-Glu-TTC", "Escherichia coli")
structure_html(svg)

# Convert to PNG
png <- structure_to_png(svg, width = 600)
```

## Session info

```{r}
#| label: session-info
sessionInfo()
```
